\texorpdfstring{Identifying Software Performance Changes Across\\ Variants and Versions}{Identifying Software Performance Changes Across Variants and Versions}

### 2020-08-24 paper.tex
over-engineered version ·but likely camera ready

	\subsubsection{Candidate Commits}
	For each configuration, we compute an approximation of the change point likelihood over commits (cf. Figure~\ref{fig:cpp_example}) to identify local maxima (peaks). We select such peaks under the condition that the peak change point likelihood is greater than a threshold:
	{\color{indigo}\begin{equation}
	\begin{gathered}
	t_\text{CPL} = \frac{1}{\vert V\vert} + N_\text{CPL} \cdot \underbrace{\sqrt{\frac{1}{\vert V\vert} \cdot \sum_{v \in V} \Big(\frac{1}{\vert V\vert} - p'(v)\Big)^2}}_{\text{standard deviation of }p'(v)}
	\label{eq:t_cpl}
	\end{gathered}
	\end{equation}
	% genauere Erklärung, was mit der STDEV gemeint/ausgedrückt wird

	The threshold $t_\text{CPL}$ is the average change point likelihood over all commits plus a factor $N_\text{CPL}$ times its corresponding standard deviation over all commits. 
	The factor $N_\text{CPL}$ (default value: 3) allows us to filter peaks that do not stand out enough and might be false positives. The greater $N_\text{CPL}$ is, the stricter the filtering of peaks. That is, for each configuration in our learning set, we obtain a set of commits (configuration-specific candidate commits) that represent possible change points.
	}To reduce variation among the obtained commits, we cluster the commits using kernel density estimation (KDE). The purpose of KDE in our setting is to estimate the probability mass function of whether a commit is a change point. The local maxima of this KDE, subsequently, represent our set of candidate commits. Note that, if two distinct change points are almost coincident commit-wise, they can be mistakenly identified by our approach as one change point that, subsequently, can be associated with configuration options belonging to the two individual change points.

@@

	For each model (for a candidate commit), we consider a configuration option as associated with a commit, if its influence $c_i$ (i.e., the absolute value of its coefficient) is greater than a threshold:% $t_\text{influence}$:
	{\color{indigo}\begin{equation}
	\begin{gathered}
	t_\text{influence} = \frac{1}{n_\text{options}} + N_\text{influence} \cdot \underbrace{\sqrt{\frac{1}{n_\text{options}} \cdot \sum \Big(\frac{1}{n_\text{options}} - c_i\Big)^2}}_{\text{standard deviation of } c_i}
	\label{eq:t_influence}
	\end{gathered}
	\end{equation}
	The parameter $N_\text{influence}$ (default value: 3) allows us to filter only estimations with low variation. If the intercept of a model exceeds this threshold, we consider this candidate commit a change point that is not configuration-specific (i.e., it affects all configurations). The outcome is a list of \emph{associations}: pairs of candidate commits and (likely) influential configuration options. }
	% unklar, warum N_influence hier eingeführt wird

@@

	We limit the total number of new measurements per iteration with a budget $n_\text{measurements}$. This budget is split between acquisition for commits and configurations with a factor $n_\text{commits\_to\_configs} = 0.5$. The measurement budget for commit acquisition is further split between exploration and exploitation with a factor $n_\text{commits\_explore} = 0.5$. That is, initially, 50\,\% of the budget are used for configuration acquisition and 25\,\% for commit exploration and exploitation, respectively. 
	{\color{indigo}As the algorithm proceeds, we multiply the factor $n_\text{commits\_to\_configs}$ by 0.95 and the factor $n_\text{commits\_explore}$ by 0.9. That is, the algorithm shifts  (1) towards sampling configurations in depth and (2) focuses on exploitation in later iterations. The rationale is that we consider pinpointing commit to configuration options a more difficult task than finding performance-changing commits.}
	% zu schnell -> genereller Shift in Richtung exloitation und weg von exploration

@@

	\subsubsection{Operationalization}
	We {\color{indigo}synthesize performance data by initializing each option with a randomly selected influence (cf. coefficients from Equation~\ref{eq:pimodel}) from the range $\left[-1,1 \right]$}. In addition, we randomly select a number of interactions among options to introduce interactions of varying degrees. The interaction degree (i.e., number of selected configuration options) follows a geometric distribution. We assign to each interaction a real-valued influence uniformly from the range $[-1, 1]$. We define six parameter ranges (cf.~Table~\ref{tab:synthetic_experiment_ranges}) for the number of configuration options $n_\text{options}$, the number of commits $n_\text{commits}$, the number of change points $n_\text{changepoints}$, the parameter for the geometric distribution of interaction degrees $p_\text{interaction}$, a factor $N_\text{initial}$, and the number of measurements per iteration $n_\text{measurements}$. The initial number of configurations is the product of $n_\text{options}$ and $N_\text{initial}$. We construct the Cartesian product of all parameter ranges and specify a maximum number of 30 iterations. In addition, we synthesize for each parameter combination in the parameter grid five different software systems by employing different seeds such that no seed is used twice. The total number of experiments with seeds is 10,\,800.% For all synthetic experiments, we specify a maximum number of 30 iterations.

@@

	For each parameter combination, we record the number of measurements at each iteration, the required number of iterations for termination, as well as each iteration's candidate solution. To assess the accuracy of a parameter combination's outcome (as well as of intermediate iterations), we use the $\text{F}_1$ score, a combination of precision and recall. \emph{Precision} refers to the fraction of correctly identified associations (true positives) among the associations of the retrieved (candidate) solution. \emph{Recall} is the fraction of total number of the relevant associations (i.e., those we intend to find). The $\text{F}_1$ score is the harmonic mean of precision ($P$) and recall ($R$), defined as $\text{F}_1 = 2\cdot\frac{P \cdot R}{P + R}$.	In our context, we defined a correctly identified change point when a commits falls in a narrow 5 commit interval from the ground truth.
	{\color{indigo}To assess efficiency, we employ the required number of iterations for termination and required measurements in relation to the $\text{F}_1$ score as proxy metrics.}
	% schwer zu verstehen

@@

	We attribute the inconclusive results for \textsc{lrzip} in part to the inconsistency of the configuration interface. For both \textsc{xz} and \textsc{lrzip}, many configuration options are alternatives. That is, for a functionality such as the compression algorithm, only one of the available options can be selected. In our approach, we employ Lasso regression for automated feature selection. This regression model might, given only a small configuration sample, have difficulties with attributing a change point to multiple alternative configuration options and rather opt for one instead.  %Across the repetitions of our experiments, we have found candidate solutions close to the two change points commits for \textsc{xz} and \textsc{oggenc}, respectively, on average, after 3 to 5 iterations. After more than five iterations, the majority of repetitions pinpointed commits correctly within our narrow 5 commit interval.
	\vspace{2mm}
	\greybox{
		\textbf{Summary ($\text{RQ}_2$)}: Our approach is able to precisely link change points to commits with real-world data. Manually associating commits with configuration options worked in many cases, whereas the automated analysis was inconclusive due to incomplete data in the repositories. We observed, at least, one inconsistency in the configuration interface of one subject system.
	}

	\section{Discussion}\label{sec:discussion}
	\paragraph{Efficiency} Among all experiments conducted with our synthetic setup, our approach yielded reasonable to high accuracy across most parameter settings. A subsequent analysis of parameter settings showed that our algorithm is not limited by the size of a problem space, but rather the number of measurements per iteration. 
	% mehr Interpretation, Details ausblenden
	{\color{indigo}Our algorithm found multiple change points across settings of up to 2,500 commits and and 2$^{64}$ configurations, resulting in a search space of about $4.6 \cdot 10^{22}$ possible measurements.
	In the context of not more than 5,000 measurements per iteration, the maximum number of measurements (after 30 iterations) added to the initial set of measurements is 150,000. For a setting with 2,500 commits, this represents around $1.4 \cdot 10^{-6}$\,\% and $3.2\cdot 10^{-16}$\,\% of possible measurements for settings with 32 and 64 options, respectively. The vast majority of cases terminated in early iterations.} Hence, we are confident that our approach scales and can efficiently approximate change points for large software systems. Notably, we are able to reproduce this finding in real-world settings when identifying the temporal location of change points. For our controlled experiment setup, we have not tuned the algorithms’ parameters to reach optimal efficiency or accuracy. However, even without tuning, we demonstrated the feasibility and efficiency of our approach. Project-specific fine tuning of several parameters is a possible extension to our approach and can further improve our results.

@@

	\section{Conclusion}
	% sperrig laut sven
	{\color{indigo}Changes in the observed performance of a software system can be configuration-specific and difficult to pinpoint to configuration options and commits. We combine both two orthogonal perspectives, the commit history and configurability of software systems to pinpoint performance changes to individual commits and configuration options. 
	We propose an iterative sampling approach that leverages similarities in performance histories of related configurations to identify performance shifts precisely and, subsequently, pinpoint them to configuration options and interactions. 
	We have shown with a synthetic data set that our approach is sound and able to identify performance change points accurately and at scale. Experiments with three configurable software systems confirm the temporal precision of our approach and have confirmed prevalent practical challenges when studying real world-software systems. It is an avenue for further research to incorporate further information, such as commit artifacts, when pinpointing performance changes.}

	\section{Acknoledgements}
	Apel’s work has been supported by the Deutsche Forschungsgemeinschaft (DFG) under the contract AP 206/11-1. Siegmund’s work has been supported by the DFG under the
	contracts SI 2171/2 and SI 2171/3-1.

>>>>

	\subsubsection{Candidate Commits}
	For each configuration, we compute an approximation of the change point likelihood over commits (cf. Figure~\ref{fig:cpp_example}) to identify local maxima (peaks). We select such peaks under the condition that the peak change point likelihood is greater than a threshold:
	\begin{equation}
	\begin{gathered}
	t_\text{CPL} = \frac{1}{\vert V\vert} + N_\text{CPL} \cdot \underbrace{\sqrt{\frac{1}{\vert V\vert} \cdot \sum_{v \in V} \Big(\frac{1}{\vert V\vert} - p'(v)\Big)^2}}_{\substack{\text{standard deviation of } p'(v) \\ \text{over all commits} }}
	\label{eq:t_cpl}
	\end{gathered}
	\end{equation}
	% genauere Erklärung, was mit der STDEV gemeint/ausgedrückt wird

	The threshold $t_\text{CPL}$ is the average change point likelihood over all commits plus a factor $N_\text{CPL}$ times its corresponding standard deviation over all commits. 
	The factor $N_\text{CPL}$ (default value: 3) allows us to filter peaks that do not stand out enough and might be false positives. The greater $N_\text{CPL}$ is, the stricter the filtering of peaks. That is, for each configuration in our learning set, we obtain a set of commits (configuration-specific candidate commits) that represent possible change points.
	To reduce variation among the obtained commits, we cluster the commits using kernel density estimation (KDE). The purpose of KDE in our setting is to estimate the probability mass function of whether a commit is a change point. The local maxima of this KDE, subsequently, represent our set of candidate commits. Note that, if two distinct change points are almost coincident commit-wise, they can be mistakenly identified by our approach as one change point that, subsequently, can be associated with configuration options belonging to the two individual change points.

@@

	For each model (for a candidate commit), we consider a configuration option as associated with a commit, if its influence $c_i$ (i.e., the absolute value of its coefficient) is greater than a threshold:% $t_\text{influence}$:
	\begin{equation}
	\begin{gathered}
	t_\text{influence} = \frac{1}{n_\text{options}} + N_\text{influence} \cdot \underbrace{\sqrt{\frac{1}{n_\text{options}} \cdot \sum \Big(\frac{1}{n_\text{options}} - c_i\Big)^2}}_{\text{standard deviation of } c_i}
	\label{eq:t_influence}
	\end{gathered}
	\end{equation}
	We consider options as influential, if their estimated influence is greater than if each configuration option were equally influential. In addition, we add to this threshold the standard deviation of the influences of all configuration options. In theory, our regression model eliminates non-influential configuration options. If more than configuration option is influential (i.e., two or more options interact) their respective estimated influence should be roughly equal. To avoid false positives in he estimation of influential configuration options, we use the standard deviation and the parameter $N_\text{influence}$ (default value: 1) to filter only estimations with low variation among the estimated influences. If the intercept of a model (i.e., the coefficient not related to any regression model variable) exceeds this threshold, we consider this candidate commit a change point that is not configuration-specific (i.e., it affects all configurations). The outcome is a list of \emph{associations}: pairs of candidate commits and (likely) influential configuration options. 
	% unklar, warum N_influence hier eingeführt wird

@@

	We limit the total number of new measurements per iteration with a budget $n_\text{measurements}$. This budget is split between acquisition for commits and configurations with a factor $n_\text{commits\_to\_configs} = 0.5$. The measurement budget for commit acquisition is further split between exploration and exploitation with a factor $n_\text{commits\_explore} = 0.5$. That is, initially, 50\,\% of the budget are used for configuration acquisition and 25\,\% for commit exploration and exploitation, respectively. 
	As the algorithm proceeds, we want to shift the budget from an equal split between exploration and exploitation towards exploitation. The rationale is that in early iterations, we focus on identifying change points in the configurations' performance histories, whereas, in later iterations, our approach focuses on pinpointing configuration options to change points. Therefore, we multiply the factor $n_\text{commits\_to\_configs}$ by 0.95 and the factor $n_\text{commits\_explore}$ by 0.9. That is, the algorithm shifts  (1) towards sampling configurations in depth and (2) focuses on exploitation in later iterations. The rationale is that we consider pinpointing commit to configuration options a more difficult task than finding performance-changing commits.
	% zu schnell -> genereller Shift in Richtung exloitation und weg von exploration

@@

	\subsubsection{Operationalization}
	We synthesize performance data by initializing each option with a randomly selected influence (cf. coefficients from Equation~\ref{eq:pimodel}) from the range $\left[-1,1 \right]$. In addition, we randomly select a number of interactions among options to introduce interactions of varying degrees. The interaction degree (i.e., number of selected configuration options) follows a geometric distribution. We assign to each interaction a real-valued influence uniformly from the range $[-1, 1]$. We define six parameter ranges (cf.~Table~\ref{tab:synthetic_experiment_ranges}) for the number of configuration options $n_\text{options}$, the number of commits $n_\text{commits}$, the number of change points $n_\text{changepoints}$, the parameter for the geometric distribution of interaction degrees $p_\text{interaction}$, a factor $N_\text{initial}$, and the number of measurements per iteration $n_\text{measurements}$. The initial number of configurations is the product of $n_\text{options}$ and $N_\text{initial}$. We construct the Cartesian product of all parameter ranges and specify a maximum number of 30 iterations. In addition, we synthesize for each parameter combination in the parameter grid five different software systems by employing different seeds such that no seed is used twice. The total number of experiments with seeds is 10,\,800.% For all synthetic experiments, we specify a maximum number of 30 iterations.

@@

	For each parameter combination, we record the number of measurements at each iteration, the required number of iterations for termination, as well as each iteration's candidate solution. To assess the accuracy of a parameter combination's outcome (as well as of intermediate iterations), we use the $\text{F}_1$ score, a combination of precision and recall. \emph{Precision} refers to the fraction of correctly identified associations (true positives) among the associations of the retrieved (candidate) solution. \emph{Recall} is the fraction of total number of the relevant associations (i.e., those we intend to find). The $\text{F}_1$ score is the harmonic mean of precision ($P$) and recall ($R$), defined as $\text{F}_1 = 2\cdot\frac{P \cdot R}{P + R}$.	In our context, we defined a correctly identified change point when a commits falls in a narrow 5 commit interval from the ground truth.
	To assess efficiency, we employ the required number of iterations for termination and required measurements in relation to the $\text{F}_1$ score as proxy metrics. The idea is that we consider an experiment run as efficient if it provides an accurate change point estimation with few iterations or few measurements.
	% schwer zu verstehen

@@

	We attribute the inconclusive results for \textsc{lrzip} in part to the inconsistency of the configuration interface. For both \textsc{xz} and \textsc{lrzip}, many configuration options are alternatives. That is, for a functionality such as the compression algorithm, only one of the available options can be selected. In our approach, we employ LASSO regression for automated feature selection. This regression model might, given only a small configuration sample, have difficulties with attributing a change point to multiple alternative configuration options and rather opt for one instead.  %Across the repetitions of our experiments, we have found candidate solutions close to the two change points commits for \textsc{xz} and \textsc{oggenc}, respectively, on average, after 3 to 5 iterations. After more than five iterations, the majority of repetitions pinpointed commits correctly within our narrow 5 commit interval.
	\vspace{2mm}
	\greybox{
		\textbf{Summary ($\text{RQ}_2$)}: Our approach is able to precisely link change points to commits with real-world data. Manually associating commits with configuration options worked in many cases, whereas the automated analysis was inconclusive due to incomplete data in the repositories. We observed, at least, one inconsistency in the configuration interface of one subject system.
	}

	\section{Discussion}\label{sec:discussion}
	\paragraph{Efficiency} Among all experiments conducted with our synthetic setup, our approach yielded reasonable to high accuracy across most parameter settings. A subsequent analysis of parameter settings showed that our algorithm is not limited by the size of a problem space, but rather the number of measurements per iteration. 
	% mehr Interpretation, Details ausblenden
	Our algorithm found multiple change points across settings of up to 2,500 commits and and 2$^{64}$ configurations, resulting in a search space of about $4.6 \cdot 10^{22}$ possible measurements.
	In the context of not more than 5,000 measurements per iteration, the maximum number of measurements (after 30 iterations) added to the initial set of measurements is 150,000. For a setting with 2,500 commits, this represents around $1.4 \cdot 10^{-6}$\,\% and $3.2\cdot 10^{-16}$\,\% of possible measurements for settings with 32 and 64 options, respectively. The vast majority of cases terminated in early iterations. Hence, we are confident that our approach scales and can efficiently approximate change points for large software systems. Notably, we are able to reproduce this finding in real-world settings when identifying the temporal location of change points. For our controlled experiment setup, we have not tuned the algorithms’ parameters to reach optimal efficiency or accuracy. However, even without tuning, we demonstrated the feasibility and efficiency of our approach. Project-specific fine tuning of several parameters is a possible extension to our approach and can further improve our results.

@@

	\section{Conclusion}
	Changes in the observed performance of a software system can be configuration-specific and difficult to pinpoint to configuration options and commits. We combine both two orthogonal perspectives, the commit history and configurability of software systems to pinpoint performance changes to individual commits and configuration options. 
	We propose an iterative sampling approach that leverages similarities in performance histories of related configurations to identify performance shifts precisely and, subsequently, pinpoint them to configuration options and interactions. 
	We have shown with a synthetic data set that our approach is sound and able to identify performance change points accurately and at scale. Experiments with three configurable software systems confirm the temporal precision of our approach and have confirmed prevalent practical challenges when studying real world-software systems. It is an avenue for further research to incorporate further information, such as commit artifacts, when pinpointing performance changes.

	\section{Acknowledgments}
	Apel’s work has been supported by the Deutsche Forschungsgemeinschaft (DFG) under the contract AP 206/11-1. Siegmund’s work has been supported by the DFG under the
	contracts SI 2171/2 and SI 2171/3-1.
